Classless Java: Tuning Java Interfaces

### 2015-12-02 paper/sections/Introduction.tex
Revisions

\item {\em Useful, general purpose methods, require special care in
  the presence of subtyping.} Methods such as $clone$, or
  \emph{fluent} setters~\cite{fowler2005fluentinterface}, not only require access to the
  internal state of an object, but they also require their types to be
  refined in subtypes.

@@

To formalize object interfaces, we propose Classless Java (CJ): a
FeatherweightJava-style~\cite{Igarashi01FJ} calculus, which captures the essence
of interfaces with default methods. The semantics of object interfaces
is given as a type-directed translation from CJ to itself. In the
resulting CJ code all object interfaces are translated into regular CJ
(and Java) interfaces with default methods. The translation is proved
to be type-preserving, ensuring that the translation does not
introduce type-errors. CJ's usefulness goes beyond serving as a
calculus to formalize object interfaces. During the development
process of CJ, we encountered a bug in the implementation of default
methods for the Eclipse Compiler for Java (ECJ). For the program revealing the 
bug, ECJ behaves differently from both our formalization and Oracle's 
Java compiler.

To evaluate the usefulness of object interfaces, we illustrate 3
applications (case studies). The first application is a simple 
solution to the Expression Problem~\cite{wadler98expression}, supporting independent 
extensibility~\cite{zenger05independentlyextensible}, and without boilerplate code. The second
application is shows how embedded DSLs using fluent interfaces~\cite{fowler2005fluentinterface} 
can be easily defined using object interfaces. The last
application is a larger case study for a simple Maze game implemented with 
multiple inheritance. For the last application we show that there is a
significant reduction in the numbers of lines of code when compared 
to an existing implementation~\cite{bono14} using plain Java 8. 
Noteworthy, all applications are implemented 
without defining a single class!

@@

\item {\bf ClassLess Java (CJ):} A simple formal calculus that models 
the essential features of Java 8 interfaces with default methods, and 
can be used to formally define the translation of object interfaces. 
We prove a type-preservation theorem, and we present a Java program 
that reveals a bug in the ECJ implementation of default methods.

>>>>

\item {\em Useful, general purpose methods, require special care in
  the presence of subtyping.} Methods such as
  \emph{fluent} setters~\cite{fowler2005fluentinterface}, not only require access to the
  internal state of an object, but they also require their return types to be
  refined in subtypes.

@@

To formalize object interfaces, we propose Classless Java (CJ): a
FeatherweightJava-style~\cite{Igarashi01FJ} calculus, which captures the essence
of interfaces with default methods. The semantics of object interfaces
is given as a syntax-directed translation from CJ to itself. In the
resulting CJ code all object interfaces are translated into regular CJ
(and Java) interfaces with default methods. The translation is proved
to be type-preserving, ensuring that the translation does not
introduce type-errors. 
%CJ's usefulness goes beyond serving as a
%calculus to formalize object interfaces. During the development
%process of CJ, we encountered a bug in the implementation of default
%methods for the Eclipse Compiler for Java (ECJ). For the program revealing the 
%bug, ECJ behaves differently from both our formalization and Oracle's 
%Java compiler.

To evaluate the usefulness of object interfaces, we illustrate 3
applications. The first application is a simple 
solution to the Expression Problem~\cite{wadler98expression}, supporting independent 
extensibility~\cite{zenger05independentlyextensible}, and without boilerplate code. The second
application is shows how embedded DSLs using fluent interfaces~\cite{fowler2005fluentinterface} 
can be easily defined using object interfaces. The last
application is a larger case study for a simple Maze game implemented with 
multiple inheritance. For the last application we show that there is a
significant reduction in the numbers of lines of code when compared 
to an existing implementation~\cite{bono14} using plain Java 8. 
Noteworthy, all applications are implemented 
without defining a single class!

@@

\item {\bf ClassLess Java (CJ):} A simple formal calculus that models 
the essential features of Java 8 interfaces with default methods, and 
can be used to formally define the translation of object interfaces. 
We prove that several properties hold for the translation, including a type-preservation.

### 2015-12-02 paper/sections/Overview.tex
Revisions

\begin{figure}
\begin{tabular}{c|c}
\includegraphics[height=4cm]{PegasusDetail.pdf}\hspace{20pt} &
\begin{minipage}{7cm}
\vspace{-90pt}
\lstinputlisting[linerange=6-17]{../UseMixinLombok/src/pegasusJava/Main.java}% APPLY:linerange=PEGASUS_JAVA
%basicstyle=\ttfamily\scriptsize
\end{minipage}
\end{tabular}
\caption{The animal system. Complete animal system on the left, and code for simplified animal system on the right. }\label{fig:pegasus}
\end{figure}

To propose a standard example, we show \Q@Animal@s with a two dimensional \Q@Point2D@ representing their \Q@location@.
Some kinds of animals are \Q@Horse@s and \Q@Bird@s.
Birds can \Q@fly@, thus their location need to be a three dimensional \Q@Point3D@.
Finally, we model \Q@Pegasus@ as a kind of \Q@Animal@ with the skills of both \Q@Horse@s and \Q@Bird@s.\footnote{
Some research argues in favour of using subtyping for modelling taxonomies, other research argue against this practice, we do not wish to take sides in this argument, but to provide an engaging example.}

\subsection{Simple Multiple Inheritance with Default Methods}

Suppose that we want to model an animal system as shown in
the left side of Figure~\ref{fig:pegasus}. \texttt{Horse} and \texttt{Bird} are subtypes
of \texttt{Animal}, with methods \texttt{run()} and \texttt{fly()},
respectively. Pegasus (one of the best known creatures in Greek
mythology) can not only \emph{run} but also \emph{fly}! This is the
place where \emph{``multiple inheritance''} is needed, because
\texttt{Pegasus} needs to obtain \texttt{fly} and \texttt{run}
functionality from both \texttt{Horse} and \texttt{Bird}.
Ignoring points, for the moment, a 
first attempt to model the animal system in Java 8 is given on the right-side 
of Figure~\ref{fig:pegasus}.
Note that the implementations of the methods \texttt{run}
and \texttt{fly} are defined inside interfaces, using default
methods. Moreover, because interfaces support multiple interface
inheritance, the interface for \texttt{Pegasus} can inherit behaviour
from both \texttt{Horse} and \texttt{Bird}. Although Java interfaces
do not allow instance fields, no form of state is needed so far to
model the animal system.  Therefore this example provides a simple
illustratation how a basic form of multiple inheritance, similar to
what can be done with traits, can be accomplished in Java 8.

\paragraph{Instantiation}
To use \texttt{Horse}, \texttt{Bird} and \texttt{Pegasus}, some
objects must be created first. A first problem with using
interfaces to model the animal system is simply that interfaces 
cannot be directly instantiated. Classes are needed for instantiation, such as:

@@

\noindent to create animal objects. Now a \texttt{Pegasus} animal can be created
using the class constructor:

@@

\noindent There is a lot of annoyance here. Firstly, the sole
purpose of classes is to provide a way to instantiate
objects. Although (in this case) it takes only one line code to
provide each of those classes, this code is essentially boilerplate
code, which does not add behaviour to the system. Secondly, 
the namespace gets filled with three additional types. For example, 
both \texttt{Horse} and \texttt{HorseImpl} are needed: \texttt{Horse}
is needed because it needs to be an inferface so that \texttt{Pegasus}
can use multiple inheritance; and \texttt{HorseImpl} is needed to
provide object instantiation. 
Note that, for this very simple animal system, plain Java 8 anonymous
classes can be used to avoid these problems.  We could have simply
instantiated Pegasus using:

@@

To model the animal system with object interfaces all that a user
needs to do is to add a \mixin annotation to the \texttt{Horse}, 
\texttt{Bird}, and \texttt{Pegasus} interfaces.  For example, 
for \texttt{Bird} and \texttt{Pegasus} the code would be as follows:

@@

The effect of the annotations is that a static \emph{factory} method called
\texttt{of} is automatically added to the interfaces. With the
\texttt{of} method a Pegasus object is instantiated as follows:

@@

\noindent Note that the generated code is transparent to a user, which
only sees the original code with the \mixin annotation. Compared to the pure
Java solution in Section~\ref{}, the solution using object interfaces
has the advantage of providing a direct mechanism for object
instantiation, which avoids adding boilerplate classes to the
namespace.

@@

The animal system modelled so far, is a simplified version of the
system presented in the left-side of Figure~\ref{fig:pegasus}, and still does not appear to 
justify the \mixin annotation. Lets now move on to modelling 
the complete animal system. In order to do this, animals will include 
a notion of points, representing their location in space. As we shall
see, modelling stateful components using plain Java 8 quickly becomes 
cumbersome. The \mixin annotation comes to rescue here and 
allows avoiding significant amounts of boilerplate code.

\paragraph{\Q@Point2D@: simple immutable data with fields}
Two dimensional points should keep track of their coordinates.  The usual
approach to model points in Java would be to use a class with fields for the
coordinates. However here we will illustrate how points are modelled with
interfaces. Since Java disallows fields inside interfaces, states are modeled
using abstract (getter) methods:

@@

\noindent Unfortunatelly, creating a new point object is quite cumbersome, even
with anonymous classes:

@@

\noindent However it is not required to use this cumbersome syntax for every
object allocation. As programmers do, for ease or reuse, the boring
repetitive code can be encapsulated in a method. A generalization of the
\texttt{of} static factory method is appropriate in this case:
\begin{lstlisting}
interface Point2D{ int x(); int y();
  static Point2D of(int x, int y){return new Point2D(){
    public int x(){return x;} public int y(){return y;}};}}
\end{lstlisting}

\vspace{-5pt}
\paragraph{\Q@Point2D@ with Object Interfaces}
This obvious constructor code can be automatically generated by our \mixin
annotation.  By annotating the interface \Q@Point2D@, a variation of the shown
static method \texttt{of} will be generated, mimicking the functionality of a
simple-minded constructor: by looking at the abstract methods it first detects
what are the fields, then generates an \Q@of@ method with one parameter for each
of them. That is, we can just write

@@

\paragraph{\texttt{with} methods in Object Interfaces}
The pattern of creating a new object by reusing most information from an old
object is very common when programming with immutable data-structures. It is
supported in our code generation as \Q@with@ methods, for example:
\begin{lstlisting}
@Obj interface Point2D { 
  int x();    int y(); // getters
  Point2D withX(int val);    Point2D withY(int val); // with methods }
\end{lstlisting}

@@

\begin{lstlisting}
interface Point2D{ int x(); int y();
  static Point2D of(int _x, int _y){return new Point2D(){
    int x=_x; int y=_y;
    public int x(){return x;}   public int y(){return y;}
    Point2D withX(int val){ return Point2D.of(val,this.y());}    
    Point2D withY(int val){ return Point2D.of(this.x(),val);}  
    };}
  Point2D withX(int val);    Point2D withY(int val);}
\end{lstlisting}
\end{comment}
Moreover, if the programmer wants a different implementation, they may 
just provide new \texttt{with} methods using \Q@default@:
\begin{lstlisting}
@Obj interface Point2D{ 
  int x(); int y();
  default Point2D withX(int val){ ... };    default Point2D withY(int val){ ... }}
\end{lstlisting}

@@

\noindent Here we declare abstract getter and setter for the mutable ``field''
\Q@location@.  Without the \mixin annotation, there is no convenient way to
instantiate \texttt{Animal}.  For \texttt{Horse}, the \mixin annotation is used
and an implementation of \texttt{run()} method is defined using \Q@default@,
which further illustrates the convenience of \emph{with} methods:

@@

\paragraph{Summary} Dealing with states (mutable or not) in object interfaces
relies on a notion of abstract state, where only methods that interact with
states are available to users. Object interfaces provide support for four
different types of abstract state operations, which are summarized in
Figure~\ref{fig:abstractstate}.  The abstract state operations are determined by
naming conventions and the types of the methods. Fluent setters are a variant of
mutable setters, and are discussed in detail in Section~\ref{sec:dsls}.

@@

\subsection{Object Interfaces and Subtyping}
\Q@Bird@s are \Q@Animal@s, but while \Q@Animal@s only need 2D
locations, \Q@Bird@s need 3D locations. Therefore when the \texttt{Bird}
interface extends the animal interface, the notion of points needs to
be refined. Such kind of refinement usually poses a challenge
in typical class-based approaches. Fortunatelly, with object interfaces,
we are able to provide a simple and effective solution.

@@

\item Define a \Q@Point2D@ field in \Q@Animal@ and define an extra \Q@int z@
  field in \Q@Bird@.  This solution is very ad-hoc, requires to basically
  duplicate the difference between \Q@Point2D@ and \Q@Point3D@ inside \Q@Bird@.
  %Again, there are many reasons this would be bad, 
  The most dramatic criticism is that it would not scale to a scenario when
  \Q@Bird@ and \Q@Point3D@ are from different programmers.

@@

\paragraph{Field type refinement with object interfaces}
Object interfaces address the challenge of type-refinement as follows:
\begin{itemize}
\item by \emph{covariant method overriding} the return type of
  \texttt{location()} is refined to \texttt{Point3D};
\item by \emph{overloading} a new setter for location is defined with a more
  precise type;
\item we provide a \Q@default@ setter with the old signature.
\end{itemize}

@@

\noindent From the type perspective, the key is the covariant method
overriding of \texttt{location()}. However from the semantic
perspective the key is the implementation for the setter with the old
signature.

\paragraph{\Q@Point3D@ and properties updater}
The \Q@Point3D@ interface is defined as follows:

@@

\noindent By using the new \Q@with@ method we may use the information for
\Q@z@ already stored in the object to forge an appropriate \Q@Point3D@
to store.  Note how all the informations about what fields sits in
\Q@Point3D@ and what in \Q@Point2D@ is properly encapsulated in the
\Q@with@ method, and is transparent for the implementer of \Q@Bird@.

\Q@with-@ methods (functionally) update a field at a time.  This can be
inefficient, and sometime hard to maintain.  Often we want to update multiple
fields simultaneously, for example using another object as source.  Following
this idea, the method \Q@with(Point2D)@ is an example of a (functional)
properties updater: it takes a certain type of object and returns a copy of the
current object with all fields that match the fields of parameter type
updated. The idea is that we want the result to be still like \Q@this@, but
modified as similar as possible to the parameter.

@@

\paragraph{Generated boilerplate}
Just to give a feeling on how much boring code \mixin is generating, we show the
generated code for the \texttt{Point3D} in Figure~\ref{fig:boilerplate}. The
generated code is very mechanical, writing such code by hand is error-prune. For
example a distracted programmer may swap the arguments of calls to
\Q@Point3D.of@.  Note how \Q@with-@ methods are automatically refined in their
return type, so that code like \Q@Point3D p=Point3D.of(1,2,3); p=p.withX(42);@
will be accepted and behave as expected. If the programmer wishes to suppress
this behaviour and keep the signature as it was, it is sufficient to redefine
the \Q@with-@ methods in the new interface repeating the old signature.  Again,
the philosophy is that if the programmer provides something directly, \mixin
does not touch it.  The cast in \Q@with(Point2D)@ is trivially safe since it is
guarded by an \Q@instanceof@ test. The idea is that if the parameter is a
subtype of the current exact type, then we can just return the parameter, as
something that is just ``more'' than \Q@this@.

\subsection{Advanced Multiple Inheritance with Object Interfaces}
Finally, we can define \texttt{Pegasus} as simple as we did in the simplified
(and stateless) version on the right-side of Figure~\ref{fig:pegasus}:

@@

\noindent Note how even the non trivial pattern for field type refinement is
transparently composed, and \texttt{Pegasus} has a \Q@Point3D@
\Q@location@.%  This works because \Q@Horse@ do not perform
% any field type refinement, otherwise we may have to choose/create a
% common subtype in order for \texttt{Pegasus} to exists.

>>>>

\begin{figure}
\begin{tabular}{c|c}
\includegraphics[height=4cm]{PegasusDetail.pdf}\hspace{20pt} &
\begin{minipage}{7cm}
\vspace{-90pt}
\lstinputlisting[linerange=6-17]{../UseMixinLombok/src/pegasusJava/Main.java}% APPLY:linerange=PEGASUS_JAVA
%basicstyle=\ttfamily\scriptsize
\end{minipage}
\end{tabular}
\caption{The animal system. Complete animal system on the left, and code for simplified animal system on the right. }\label{fig:pegasus}
\end{figure}
\bruno{Yanlin: figure needs to rename point to location}

This section illustrates how the \mixin annotation enables expressive
idioms using multiple inheritance. 
To propose a standard example for multiple inheritance, we show \Q@Animal@s with a two dimensional \Q@Point2D@ representing their \Q@location@.
Some kinds of animals are \Q@Horse@s and \Q@Bird@s.
Birds can \Q@fly@, thus their location need to be a three dimensional \Q@Point3D@.
Finally, we model \Q@Pegasus@ (one of the best known creatures in Greek
mythology) as a kind of \Q@Animal@ with the skills
of both \Q@Horse@s and \Q@Bird@s. A simple class diagram illustrating
the basic system in given in the left side of Figure~\ref{fig:pegasus}.\footnote{
Some research argues in favour of using subtyping for modelling taxonomies, other research argue against this practice, we do not wish to take sides in this argument, but to provide an engaging example.}

\subsection{Simple Multiple Inheritance with Default
  Methods}\label{sec:simple}

Before we model the final animal system, we first start with a
simplified version without locations. This version serves the purpose of illustrating how 
Java 8 default methods can already model simple forms of multiple inheritance.
\texttt{Horse} and \texttt{Bird} are subtypes
of \texttt{Animal}, with methods \texttt{run()} and \texttt{fly()},
respectively. Pegasus can not only \emph{run} but also \emph{fly}! This is the
place where \emph{``multiple inheritance''} is needed, because
\texttt{Pegasus} needs to obtain \texttt{fly} and \texttt{run}
functionality from both \texttt{Horse} and \texttt{Bird}.
A first attempt to model the animal system in Java 8 is given on the right-side 
of Figure~\ref{fig:pegasus}.
Note that the implementations of the methods \texttt{run}
and \texttt{fly} are defined inside interfaces, using default
methods. Moreover, because interfaces support multiple interface
inheritance, the interface for \texttt{Pegasus} can inherit behaviour
from both \texttt{Horse} and \texttt{Bird}. Although Java interfaces
do not allow instance fields, no form of state is needed so far to
model the animal system.

\paragraph{Instantiation}
To use \texttt{Horse}, \texttt{Bird} and \texttt{Pegasus}, some
objects must be created first. A first problem with using
interfaces to model the animal system is simply that interfaces 
cannot be directly instantiated. Classes, such as:

@@

\noindent are needed for instantiation. Now a \texttt{Pegasus} animal can be created
using the class constructor:

@@

\noindent There is a some annoyances here. Firstly, the sole
purpose of classes is to provide a way to instantiate
objects. Although (in this case) it takes only one line code to
provide each of those classes, this code is essentially boilerplate
code, which does not add behaviour to the system. Secondly, 
the namespace gets filled with three additional types. For example, 
both \texttt{Horse} and \texttt{HorseImpl} are needed: \texttt{Horse}
is needed because it needs to be an inferface so that \texttt{Pegasus}
can use multiple inheritance; and \texttt{HorseImpl} is needed to
provide object instantiation. 
Note that, for this very simple animal system, plain Java 8 anonymous
classes can be used to avoid these problems.  We could have simply
instantiated Pegasus using:

@@

To model the animal system with object interfaces all that a user
needs to do is to add a \mixin annotation to the \texttt{Horse}, 
\texttt{Bird}, and \texttt{Pegasus} interfaces:

@@

\noindent The effect of the annotations is that a static \emph{factory} method called
\texttt{of} is automatically added to the interfaces. With the
\texttt{of} method a Pegasus object is instantiated as follows:

@@

\noindent Note that the generated code is transparent to a user, which
only sees the original code with the \mixin annotation. Compared to the pure
Java solution in Section~\ref{sec:simple}, the solution using object interfaces
has the advantage of providing a direct mechanism for object
instantiation, which avoids adding boilerplate classes to the
namespace.

@@

The animal system modelled so far is a simplified version of the
system presented in the left-side of Figure~\ref{fig:pegasus}, and still does not appear to 
justify the \mixin annotation. Lets now move on to modelling 
the complete animal system. In order to do this animals will include 
a notion of points, representing their location in space. As we shall
see, modelling stateful components using plain Java 8 quickly becomes 
cumbersome. The \mixin annotation comes to rescue here and 
allows avoiding significant amounts of boilerplate code.

\paragraph{\Q@Point2D@: simple immutable data with fields}
Two dimensional points should keep track of their coordinates.  The usual
approach to model points in Java would be to use a class with fields for the
coordinates. However here we will illustrate how points are modelled with
interfaces. Since Java disallows fields inside interfaces, state is modeled
using abstract (getter) methods:

@@

\noindent Unfortunatelly, creating a new point object is cumbersome, even
with anonymous classes:

@@

\noindent However this cumbersome syntax is not required for every
object allocation. As programmers do, for ease or reuse, the boring
repetitive code can be encapsulated in a method. A generalization of the
\texttt{of} static factory method is appropriate in this case:
\begin{lstlisting}
interface Point2D{ int x(); int y();
  static Point2D of(int x, int y){return new Point2D(){
    public int x(){return x;} public int y(){return y;}};}}
\end{lstlisting}

\vspace{-5pt}
\paragraph{\Q@Point2D@ with Object Interfaces}
This obvious ``constructor'' code can be automatically generated by our \mixin
annotation.  By annotating the interface \Q@Point2D@, a variation of the shown
static method \texttt{of} will be generated, mimicking the functionality of a
simple-minded constructor. \mixin first looks at the abstract methods and detects
what are the fields, then generates an \Q@of@ method with one parameter for each
of them. That is, we can just write

@@

\paragraph{\texttt{wither} methods in Object Interfaces}
The pattern of creating a new object by reusing most information from an old
object is very common when programming with immutable
data-structures. As such, it is
supported in our code generation as \Q@with@ methods. For example:
\begin{lstlisting}
@Obj interface Point2D { 
  int x();    int y(); // getters
  Point2D withX(int val);    Point2D withY(int val); // with methods }
\end{lstlisting}

@@

\begin{lstlisting}
interface Point2D{ int x(); int y();
  static Point2D of(int _x, int _y){return new Point2D(){
    int x=_x; int y=_y;
    public int x(){return x;}   public int y(){return y;}
    Point2D withX(int val){ return Point2D.of(val,this.y());}    
    Point2D withY(int val){ return Point2D.of(this.x(),val);}  
    };}
  Point2D withX(int val);    Point2D withY(int val);}
\end{lstlisting}
\end{comment}
Moreover, if the programmer wants a different implementation, he may 
provide an alternative implementation using \Q@default@ methods:
\begin{lstlisting}
@Obj interface Point2D{ 
  int x(); int y();
  default Point2D withX(int val){ ... };    default Point2D withY(int val){ ... }}
\end{lstlisting}

@@

\noindent Here we declare abstract getter and setter for the mutable ``field''
\Q@location@.  Without the \mixin annotation, there is no convenient way to
instantiate \texttt{Animal}.  For \texttt{Horse}, the \mixin annotation is used
and an implementation of \texttt{run()} is defined using a \Q@default@
method. The implementation of \texttt{run()} further illustrates the convenience of \emph{with} methods:

@@

\paragraph{Summary} Dealing with state (mutable or not) in object interfaces
relies on a notion of abstract state, where only methods that interact with
state are available to users. Object interfaces provide support for four
different types of abstract state operations, which are summarized in
Figure~\ref{fig:abstractstate}.  The abstract state operations are determined by
naming conventions and the types of the methods. Fluent setters are a variant of
mutable setters, and are discussed in detail in Section~\ref{sec:dsls}.

@@

\subsection{Object Interfaces and Subtyping}
\Q@Bird@s are \Q@Animal@s, but while \Q@Animal@s only need 2D
locations, \Q@Bird@s need 3D locations. Therefore when the \texttt{Bird}
interface extends the \Q@Animal@ interface, the notion of points needs to
be refined. Such kind of refinement usually poses a challenge
in typical class-based approaches. Fortunatelly, with object interfaces,
we are able to provide a simple and effective solution.

@@

\item Define a \Q@Point2D@ field in \Q@Animal@ and define an extra \Q@int z@
  field in \Q@Bird@.  This solution is very ad-hoc, requiring to basically
  duplicate the difference between \Q@Point2D@ and \Q@Point3D@ inside \Q@Bird@.
  %Again, there are many reasons this would be bad, 
  The most dramatic criticism is that it would not scale to a scenario when
  \Q@Bird@ and \Q@Point3D@ are from different programmers.

@@

\paragraph{Field type refinement with object interfaces}
Object interfaces address the challenge of type-refinement as follows:
\begin{itemize}
\item by \emph{covariant method overriding} the return type of
  \texttt{location()} is refined to \texttt{Point3D};
\item by \emph{overloading} a new setter for location is defined with a more
  precise type;
\item a \Q@default@ setter implementation with the old signature is provided.
\end{itemize}

Thus, with object interfaces, the code for the \Q@Bird@ interface is:

@@

\noindent From the type perspective, the key is the covariant method
overriding of \texttt{location()}. However from the semantic
perspective the key is the implementation for the setter with the old
signature (\Q@location(Point2D)@). The key to the setter
implementation is a new type of \@with@ method, called a property updater.

\paragraph{\Q@Point3D@ and properties updaters}
The \Q@Point3D@ interface is defined as follows:

@@

\noindent Note that this interface includes a different type of
\Q@with@ method, which takes a \Q@Point2D@ as an argument.
The old wither methods (such as \Q@withX@) functionally update a field at a time.  This can be
inefficient, and sometimes hard to maintain.  Often we want to update multiple
fields simultaneously, for example using another object as source.  Following
this idea, the method \Q@with(Point2D)@ is an example of a (functional)
properties updater: it takes a certain type of object and returns a copy of the
current object with all fields that match the fields of parameter type
updated. The idea is that we want the result to still be like \Q@this@, but
modified to be as similar as possible to the parameter.

With the new \Q@with@ method we may use the information for
\Q@z@ already stored in the object to forge an appropriate \Q@Point3D@
to store. Note how all the information about what fields sit in
\Q@Point3D@ and in \Q@Point2D@ is properly encapsulated in the
\Q@with@ method, and is transparent to the implementer of \Q@Bird@.

@@

\paragraph{Generated boilerplate}
Just to give a feeling on how much boring code \mixin is generating, we show the
generated code for the \texttt{Point3D} in Figure~\ref{fig:boilerplate}. The
generated code is very mechanical, writing such code by hand is error-prone. For
example a distracted programmer may swap the arguments of calls to
\Q@Point3D.of@.  Note how \Q@with-@ methods are automatically refined in their
return type, so that code like:

\begin{lstlisting}
Point3D p=Point3D.of(1,2,3); p=p.withX(42);
\end{lstlisting}

\noindent will be accepted and behave as expected. If the programmer wishes to suppress
this behaviour and keep the signature as it was, it is sufficient to redefine
the \Q@with-@ methods in the new interface repeating the old signature.  Again,
the philosophy is that if the programmer provides something directly, \mixin
does not touch it.  The cast in \Q@with(Point2D)@ is trivially safe since it is
guarded by an \Q@instanceof@ test. The idea is that if the parameter is a
subtype of the current exact type, then we can just return the parameter, as
something that is just ``more'' than \Q@this@.

\subsection{Advanced Multiple Inheritance with Object Interfaces}
Finally, we can define Pegasus as simply as we did in the simplified (and stateless) 
version in the right-side of Figure~\ref{fig:pegasus}:

@@

\noindent Note how even the non trivial pattern for field type refinement is
transparently composed, and \texttt{Pegasus} has a \Q@Point3D@
\Q@location@. This works because \Q@Horse@ do not perform
any field type refinement, otherwise we may have to choose/create a
common subtype in order for \texttt{Pegasus} to exists.
